---
title: "Host DGE BY04"
author: "BY"
date: "2024-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library("tximport")
library("readr")
library("DESeq2")
library("ggplot2")
library("gplots")
library("RColorBrewer")
library("PoiClaClu")
library("pheatmap")
library("IHW")
library("ashr")
library("tidyverse")
library(ggrepel)
```

Specify working directory and load sample data from salmon output
```{r}
setwd("~/Lucinid_project/BY04/DGE_DESeq2_val_salmonv0.14")

#upload sample/condition table
samples <- read.table("samples.txt", header=TRUE) 
#specify the path to the quantfiles using the appropriate names in "samples"
files <- file.path(samples$run, "quant.sf")
#i think this labels the paths with sample names
names(files) <- samples$run 
```

import gene to transcript mapping file
```{r}
#upload transcript to gene mapping table
tx2gene <- read.csv("./ref_assembly_corset.gene_trans_map.csv")

# import quantification data for DESeq2 using the tximport function
txi <- tximport(files, type="salmon", tx2gene=tx2gene, dropInfReps=TRUE) 

```

Construct a DESeqDataSet
```{r}
#construct a DESeqDataSet from the txi object and sample information in samples
ddsTxi <- DESeqDataSetFromTximport(txi,
                                   colData = samples,
                                   design = ~ condition)

#remove genes with low expression values
keep <- rowSums(counts(ddsTxi)) >= 10
dds <- ddsTxi[keep,]
#check levels of conditions
levels(dds$condition)
```

#####Exploratory visualisation####
Transform counts
```{r}
#These steps transform the counts matrix in different ways
#use blind=FALSE only if DESeq function has been run. 
vsd <- vst(dds, blind=TRUE)
rld <- rlog(dds, blind=TRUE)
ntd <- normTransform(dds)
```

Basic PCAs
```{r}
#basic PCA
#plotPCA(vsd, intgroup=c("condition"))

#This one is for the rld transformed counts
(data <- plotPCA(rld, intgroup = c( "run", "condition"), returnData=TRUE)) # return the data used for plotting
percentVar <- round(100 * attr(data, "percentVar"))
qplot(PC1, PC2, color=run, shape=condition, data=data) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance"))

#This one is for the vsd transformed counts
(PCAdata <- plotPCA(vsd, intgroup = c( "batch", "condition"), returnData=TRUE)) # return the data used for plotting
percentVar_vsd <- round(100 * attr(PCAdata, "percentVar"))
qplot(PC1, PC2, color=run, shape=condition, data=PCAdata) +
  xlab(paste0("PC1: ",percentVar_vsd[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_vsd[2],"% variance"))
#https://support.bioconductor.org/p/90791/


### Alternative PCA for vsd counts
pcaData_vsd <- plotPCA(vsd, intgroup=c("batch", "condition"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData_vsd, "percentVar"))

# Create PCA plot
PCA_plot_vsd <- ggplot(pcaData_vsd, aes(PC1, PC2, color=condition, label=rownames(pcaData_vsd))) +
  geom_point(size=2) +
  geom_text_repel(size=2, hjust=0, vjust=0.5) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed()

# Print the plot
PCA_plot_vsd
```



Measuring euclidean distance between samples
```{r}
#calculate sample distances
sampleDists <- dist( t( assay(rld) ) )
sampleDists

#Plot sample euclidean distances heatmap
sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste( rld$run, rld$condition, sep="-" )
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)#Either colors or colors 2 would do
colors2 <- colorRampPalette( rev(brewer.pal(9, "RdBu")) )(255)
hc <- hclust(sampleDists) # provides hierarchical clustering to hc 
heatmap.2( sampleDistMatrix, Rowv=as.dendrogram(hc),
           symm=TRUE, trace="none", col=colors,
           margins=c(2,10), labCol=FALSE )
```

NDMS
```{r}
#MDS plot
mds <- as.data.frame(colData(vsd))  %>%
  cbind(cmdscale(sampleDistMatrix))
ggplot(mds, aes(x = `1`, y = `2`, color = condition, shape = condition)) +
  geom_point(size = 3) + coord_fixed()
```

Calculate and plot Poisson Distance of samples
```{r}
#Calculate Poisson Distance of samples
#requires PoiClaClu
poisd <- PoissonDistance(t(counts(dds)))

#Plot sample poisson distance heatmap
samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- paste( dds$run, dds$condition, sep="-" )
colors <- colorRampPalette( rev(brewer.pal(9, "RdBu")) )(255)
hcpoisson <- hclust(poisd$dd) # provides hierarchical clustering to hcpoisson
heatmap.2( samplePoisDistMatrix, Rowv=as.dendrogram(hcpoisson),
           symm=TRUE, trace="none", col=colors,
           margins=c(2,10), labCol=FALSE )
```



####Differential expression analyses start here####
Run DESeq
```{r}
levels(dds$condition) #check order of levels

#the estimation of size factors (which control for differences in the library size of the sequencing experiments), 
#the estimation of dispersion for each gene, and fitting a generalized linear model.
dds <- DESeq(dds) #only one condition so this is not a multifactorial design


```

Pairwise comparisons
#test Independent hypothesis weighting for p-value adjustment. It is important to put threshold in because it otherwise tests for genes that are not zero
#requires library("IHW") for independent hypothesis testing and library("ashr") for shrinkage
#contrast = c('factorName','numeratorLevel','denominatorLevel')
#lfcthreshold of 1 = minumum fold change of 2, threshold of 0.58 means minimum fold change of 1.5
```{r}
##Fresh vs Sulphidic
#set the lfcthreshold to 0.58 (1.5 fold change) and false discovery rate threshold to 0.05 (5% false positives)
resFS_IHW_fc1.5 <- results(dds, alpha = 0.05, lfcThreshold = 0.58, altHypothesis="greaterAbs", filterFun=ihw, contrast=c("condition","Fresh","Sulphidic"))
resFS_IHW_fc1.5_shrunk <- lfcShrink(dds, type = c("ashr"), contrast=c("condition","Fresh","Sulphidic"), res=resFS_IHW_fc1.5)

resFS_IHW_fc1.5_shrunkSig <- subset(resFS_IHW_fc1.5_shrunk, padj < 0.05)
resFS_IHW_fc1.5_shrunkSig_ordered <- resFS_IHW_fc1.5_shrunkSig[order(resFS_IHW_fc1.5_shrunkSig$log2FoldChange, decreasing=TRUE),] # Sort it by the log2 fold change estimate to get the significant genes with the strongest upregulation
resFS_IHW_fc1.5_shrunkSig_ordered_up <- subset(resFS_IHW_fc1.5_shrunkSig_ordered, log2FoldChange > 0)
resFS_IHW_fc1.5_shrunkSig_ordered_down <- subset(resFS_IHW_fc1.5_shrunkSig_ordered, log2FoldChange < 0)
write.table(resFS_IHW_fc1.5_shrunkSig_ordered, file="resFS_IHW_fc1.5_shrunkSig_ordered.tsv", sep = "\t")
write.table(resFS_IHW_fc1.5_shrunkSig_ordered_up, file="resFS_IHW_fc1.5_shrunkSig_ordered_up.tsv", sep = "\t")
write.table(resFS_IHW_fc1.5_shrunkSig_ordered_down, file="resFS_IHW_fc1.5_shrunkSig_ordered_down.tsv", sep = "\t")

```

```{r}
##Fresh vs Washed


#lower the lfcthreshold 0.58 (1.5 fold change) and lower the false discovery rate threshold to 0.05 (5% false positives)
resFW_IHW_fc1.5 <- results(dds, alpha = 0.05, lfcThreshold = 0.58, altHypothesis="greaterAbs", filterFun=ihw, contrast=c("condition","Fresh","Washed"))
resFW_IHW_fc1.5_shrunk <- lfcShrink(dds, type = c("ashr"), contrast=c("condition","Fresh","Washed"), res=resFW_IHW_fc1.5)
resFW_IHW_fc1.5_shrunkSig <- subset(resFW_IHW_fc1.5_shrunk, padj < 0.05)
resFW_IHW_fc1.5_shrunkSig_ordered <- resFW_IHW_fc1.5_shrunkSig[order(resFW_IHW_fc1.5_shrunkSig$log2FoldChange, decreasing=TRUE),] # Sort it by the log2 fold change estimate to get the significant genes with the strongest upregulation
resFW_IHW_fc1.5_shrunkSig_ordered_up <- subset(resFW_IHW_fc1.5_shrunkSig_ordered, log2FoldChange > 0)
resFW_IHW_fc1.5_shrunkSig_ordered_down <- subset(resFW_IHW_fc1.5_shrunkSig_ordered, log2FoldChange < 0)
write.table(resFW_IHW_fc1.5_shrunkSig_ordered, file="resFW_IHW_fc1.5_shrunkSig_ordered.tsv", sep = "\t")
write.table(resFW_IHW_fc1.5_shrunkSig_ordered_up, file="resFW_IHW_fc1.5_shrunkSig_ordered_up.tsv", sep = "\t")
write.table(resFW_IHW_fc1.5_shrunkSig_ordered_down, file="resFW_IHW_fc1.5_shrunkSig_ordered_down.tsv", sep = "\t")
```

```{r}
##Sulphidic vs Washed

#lower lfcthreshold and alpha
resSW_IHW_fc1.5 <- results(dds, alpha = 0.05, lfcThreshold = 0.58, altHypothesis="greaterAbs", filterFun=ihw, contrast=c("condition","Sulphidic","Washed"))

resSW_IHW_fc1.5_shrunk <- lfcShrink(dds, type = c("ashr"), contrast=c("condition","Sulphidic","Washed"), res=resFW_IHW_fc1.5)
resSW_IHW_fc1.5_shrunkSig <- subset(resSW_IHW_fc1.5_shrunk, padj < 0.05)
resSW_IHW_fc1.5_shrunkSig_ordered <- resSW_IHW_fc1.5_shrunkSig[order(resSW_IHW_fc1.5_shrunkSig$log2FoldChange, decreasing=TRUE),] # Sort it by the log2 fold change estimate to get the significant genes with the strongest upregulation
resSW_IHW_fc1.5_shrunkSig_ordered_up <- subset(resSW_IHW_fc1.5_shrunkSig_ordered, log2FoldChange > 0)
resSW_IHW_fc1.5_shrunkSig_ordered_down <- subset(resSW_IHW_fc1.5_shrunkSig_ordered, log2FoldChange < 0)
write.table(resSW_IHW_fc1.5_shrunkSig_ordered, file="resSW_IHW_fc1.5_shrunkSig_ordered.tsv", sep = "\t")
write.table(resSW_IHW_fc1.5_shrunkSig_ordered_up, file="resSW_IHW_fc1.5_shrunkSig_ordered_up.tsv", sep = "\t")
write.table(resSW_IHW_fc1.5_shrunkSig_ordered_down, file="resSW_IHW_fc1.5_shrunkSig_ordered_down.tsv", sep = "\t")

```

```{r}
#Write and export a table of vst counts
vsd_deseq <- vst(dds, blind=FALSE)
write.table(assay(vsd_deseq), file="vstBY04.txt", sep="\t")

#Write and export a table of un-normalised untransformed counts
raw <- counts(dds, normalized = FALSE)
write.csv(raw, file="raw_counts.csv")
```

```{r}
#explore PCAs
library(PCAtools)
#convert vsd to matrix that can be used
vst <- assay(vst(dds))
#run PCA
p <- pca(vst, metadata = colData(dds), removeVar = 0.1)
#create a scree plot
screeplot(p, axisLabSize = 18, titleLabSize = 22, hline = 80,    returnPlot = FALSE) +
    geom_label(aes(6, 80, label = '80% explained variation', vjust = -1, size = 8))


#make a biplot:representation of samples in a 2-dimensional space, usually focusing on just the first two PCs:
biplot(p,
       # loadings parameters
       showLoadings = TRUE,
       lengthLoadingsArrowsFactor = 1.5,
       sizeLoadingsNames = 4,
       colLoadingsNames = 'red4',
       # other parameters
       lab = NULL,
       colby = 'condition', 
       hline = 0, vline = c(-25, 0, 25),
       vlineType = c('dotdash', 'solid', 'dashed'),
       gridlines.major = FALSE, gridlines.minor = FALSE,
       pointSize = 5,
       legendPosition = 'left', legendLabSize = 14, legendIconSize = 8.0,
       drawConnectors = FALSE,
       title = 'PCA bi-plot',
       subtitle = 'PC1 versus PC2',
       caption = '11 PCs ≈ 80%')


#make a pairs plot
 pairsplot(p, colby = 'condition') 
 pairsplot(p,
    components = getComponents(p, c(1:3)),
    triangle = TRUE, trianglelabSize = 12,
    hline = 0, vline = 0,
    pointSize = 0.4,
    gridlines.major = FALSE, gridlines.minor = FALSE,
    colby = 'condition',
    title = 'Pairs plot', plotaxes = FALSE,
    margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
 
 #plot the loadings, should match the top hits
   plotloadings(p,
    rangeRetain = 0.01,
    labSize = 4.0,
    title = 'Loadings plot',
    subtitle = 'PC1, PC2, PC3, PC4, PC5',
    caption = 'Top 1% variables',
    shape = 24,
    col = c('limegreen', 'black', 'red3'),
    drawConnectors = TRUE)
```