---
title: "WGCNA_4_BY04_host-var20_relateSignedModsToExt_13082024"
author: "BY"
date: "2024-08-13
"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

setup work space and load data from previous step
```{r}
#=====================================================================================
#
#  Code chunk 1
#
#=====================================================================================

# Load the WGCNA package
library(WGCNA)
# The following setting is important, do not omit.
options(stringsAsFactors = FALSE);
# Load the expression and trait data saved in the first part
lnames = load(file = "WGCNA_2_BY04-host-var20-dataInput_13082024.RData");
#The variable lnames contains the names of loaded variables.
lnames
# Load network data saved in the second part.
lnames = load(file = "WGCNA_3_BY04_host-var20_SignedNetworkConstr_summary_13082024.RData");
lnames

```


Do module and trait correlation analysis
```{r}
#=====================================================================================
#
#  Code chunk 2
#
#=====================================================================================
#correlate eigengenes with external traits and look for the most significant associations

# Define numbers of genes and samples
nGenes = ncol(datExpr);
nSamples = nrow(datExpr);
# Recalculate MEs with color labels
MEs0 = moduleEigengenes(datExpr, moduleColors)$eigengenes
MEs = orderMEs(MEs0, greyLast= FALSE)
table(moduleColors)

#remove grey module because it contains genes that have not been clustered in any module
MEs = removeGreyME(MEs, greyMEName = paste(moduleColor.getMEprefix(), "grey", sep=""))

datTraits1 <- datTraits[,3:9] #keep only the Condition traits
datTraits2 <- datTraits1[c(1,2,4,3,5,6,7)] #re-order the traits

moduleTraitCor = cor(MEs, datTraits2, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);
```

```{r}
# Build a tree of the MEs - This should match order of MEs in heatmap as long as greyLast is FALSE
MEDiss = 1-cor(MEs);
METree = hclust(as.dist(MEDiss), method = "average");
sizeGrWindow(7, 6)
plot(METree, main = "Clustering of module eigengenes", xlab = "", sub = "")


pdf(file = "WGCNA_4_BY04_host-var20_relateSignedModsToExt_13082024_moduleclustering.pdf", width = 10, height = 8)
plot(METree, main = "Clustering of module eigengenes", xlab = "", sub = "")

dev.off()

```


#Heatmap for graphical visualisation of traits and associated correlation values
```{r}
#=====================================================================================
#
#  Code chunk 3
#
#=====================================================================================

#Graphical visualisation of traits and associated correlation values

sizeGrWindow(12,8)
# Will display correlations and their p-values
textMatrix =  paste(signif(moduleTraitCor, 2), "\n(",
                    signif(moduleTraitPvalue, 1), ")", sep = "");
dim(textMatrix) = dim(moduleTraitCor)
par(mar = c(6, 8.5, 4, 6));
```

```{r}
# Display the correlation values within a heatmap plot

library("RColorBrewer")
hmcol<- colorRampPalette(c("#79b6c9", "white", "#e27a37"))(100)
par(mar = c(6, 8, 6, 6));
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(datTraits2),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = hmcol,
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Host Module-trait relationships"))

pdf(file = "WGCNA_4_BY04_host-var20_relateSignedModsToExt_13082024_module_trait_rs.pdf", width = 8, height = 8)
par(mar = c(6, 8, 6, 6));
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(datTraits2),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = hmcol,
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Host Module-trait relationships"))
dev.off()

```

```{r}
####Another way to plot MEs in a heatmap####

library(pheatmap)

pheatmap(as.matrix(moduleTraitCor), col=hmcol, cluster_cols=FALSE, cluster_rows=T, show_rownames=T, clustering_method = "median", border_color=FALSE, scale = "none", display_numbers = TRUE)

```


Quantify associations of individual genes with a trait of interest (Condition) by defining Gene Significance - must be done individually for each condition of interest
```{r}
#=====================================================================================
#
#  Code chunk 4.1 - ConditionSW
#
#=====================================================================================
#Gene relationship to trait and important modules. 
#This has to be done separately for each trait of interest
#quantify associations of individual genes with our trait of interest (Condition) by defining Gene Significance


# Define variable Condition containing the Condition column of datTrait
ConditionSW = as.data.frame(datTraits$Condition.S.vs.W);
names(ConditionSW) = "ConditionSW"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificanceSW = as.data.frame(cor(datExpr, ConditionSW, use = "p"));
GSPvalueSW = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificanceSW), nSamples));

names(geneTraitSignificanceSW) = paste("GS.", names(ConditionSW), sep="");
names(GSPvalueSW) = paste("p.GS.", names(ConditionSW), sep="");

```

```{r}
#=====================================================================================
#
#  Code chunk 4.2 - ConditionFW
#
#=====================================================================================
#Gene relationship to trait and important modules. 
#This has to be done separately for each trait of interest
#quantify associations of individual genes with our trait of interest (Condition) by defining Gene Significance


# Define variable Condition containing the Condition column of datTrait
ConditionFW = as.data.frame(datTraits$Condition.F.vs.W);
names(ConditionFW) = "ConditionFW"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificanceFW = as.data.frame(cor(datExpr, ConditionFW, use = "p"));
GSPvalueFW = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificanceFW), nSamples));

names(geneTraitSignificanceFW) = paste("GS.", names(ConditionFW), sep="");
names(GSPvalueFW) = paste("p.GS.", names(ConditionFW), sep="");
```

```{r}
#=====================================================================================
#
#  Code chunk 4.3 - WvAll
#
#=====================================================================================
#Gene relationship to trait and important modules. 
#This has to be done separately for each trait of interest
#quantify associations of individual genes with our trait of interest (Condition) by defining Gene Significance

# Define variable Condition containing the Condition column of datTrait
ConditionWvAll = as.data.frame(datTraits$Condition.W.vs.all);
names(ConditionWvAll) = "ConditionWvAll"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificanceWvAll = as.data.frame(cor(datExpr, ConditionWvAll, use = "p"));
GSPvalueWvAll = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificanceWvAll), nSamples));

names(geneTraitSignificanceWvAll) = paste("GS.", names(ConditionWvAll), sep="");
names(GSPvalueWvAll) = paste("p.GS.", names(ConditionWvAll), sep="");
```

```{r}
#=====================================================================================
#
#  Code chunk 4.4 - FvAll
#
#=====================================================================================
#Gene relationship to trait and important modules. 
#This has to be done separately for each trait of interest
#quantify associations of individual genes with our trait of interest (Condition) by defining Gene Significance

# Define variable Condition containing the Condition column of datTrait
ConditionFvAll = as.data.frame(datTraits$Condition.F.vs.all);
names(ConditionFvAll) = "ConditionFvAll"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificanceFvAll = as.data.frame(cor(datExpr, ConditionFvAll, use = "p"));
GSPvalueFvAll = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificanceFvAll), nSamples));

names(geneTraitSignificanceFvAll) = paste("GS.", names(ConditionFvAll), sep="");
names(GSPvalueFvAll) = paste("p.GS.", names(ConditionFvAll), sep="");
```

```{r}
#=====================================================================================
#
#  Code chunk 4.5 - SvAll
#
#=====================================================================================
#Gene relationship to trait and important modules. 
#This has to be done separately for each trait of interest
#quantify associations of individual genes with our trait of interest (Condition) by defining Gene Significance

# Define variable Condition containing the Condition column of datTrait
ConditionSvAll = as.data.frame(datTraits$Condition.S.vs.all);
names(ConditionSvAll) = "ConditionSvAll"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificanceSvAll = as.data.frame(cor(datExpr, ConditionSvAll, use = "p"));
GSPvalueSvAll = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificanceSvAll), nSamples));

names(geneTraitSignificanceSvAll) = paste("GS.", names(ConditionSvAll), sep="");
names(GSPvalueSvAll) = paste("p.GS.", names(ConditionSvAll), sep="");

```

```{r}
#=====================================================================================
#
#  Code chunk 4.6 - SoxB
#
#=====================================================================================
#Gene relationship to trait and important modules. 
#This has to be done separately for each trait of interest
#quantify associations of individual genes with our trait of interest (Condition) by defining Gene Significance


# Define variable Condition containing the Condition column of datTrait
ConditionSoxB = as.data.frame(datTraits$SoxB);
names(ConditionSoxB) = "ConditionSoxB"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificanceSoxB = as.data.frame(cor(datExpr, ConditionSoxB, use = "p"));
GSPvalueSoxB = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificanceSoxB), nSamples));

names(geneTraitSignificanceSoxB) = paste("GS.", names(ConditionSoxB), sep="");
names(GSPvalueSoxB) = paste("p.GS.", names(ConditionSoxB), sep="");

```

```{r}
#=====================================================================================
#
#  Code chunk 4.7 - ConditionFS
#
#=====================================================================================
#Gene relationship to trait and important modules. 
#This has to be done separately for each trait of interest
#quantify associations of individual genes with our trait of interest (Condition) by defining Gene Significance


# Define variable Condition containing the Condition column of datTrait
ConditionFS = as.data.frame(datTraits$Condition.F.vs.S);
names(ConditionFS) = "ConditionFS"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificanceFS = as.data.frame(cor(datExpr, ConditionFS, use = "p"));
GSPvalueFS = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificanceFS), nSamples));

names(geneTraitSignificanceFS) = paste("GS.", names(ConditionFS), sep="");
names(GSPvalueFS) = paste("p.GS.", names(ConditionFS), sep="");
```

```{r}
#=====================================================================================
#
#  Code chunk 4.8 - ConditionAQvAll
#
#=====================================================================================
#Gene relationship to trait and important modules. 
#This has to be done separately for each trait of interest
#quantify associations of individual genes with our trait of interest (Condition) by defining Gene Significance


# Define variable Condition containing the Condition column of datTrait
ConditionAQvAll = as.data.frame(datTraits$Condition.AQ.vs.all);
names(ConditionAQvAll) = "ConditionAQvAll"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificanceAQvAll = as.data.frame(cor(datExpr, ConditionAQvAll, use = "p"));
GSPvalueAQvAll = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificanceAQvAll), nSamples));

names(geneTraitSignificanceAQvAll) = paste("GS.", names(ConditionAQvAll), sep="");
names(GSPvalueAQvAll) = paste("p.GS.", names(ConditionAQvAll), sep="");
```

Intramodular analysis: identifying genes with high GS and MM
```{r}
#=====================================================================================
#
#  Code chunk 5
#
#=====================================================================================
#Intramodular analysis: identifying genes with high GS and MM

module = "blue"
column = match(module, modNames);
moduleGenes = moduleColors==module;

sizeGrWindow(7, 7);
par(mfrow = c(1,1));
verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                   abs(geneTraitSignificanceSvAll[moduleGenes, 1]),
                   xlab = paste("Module Membership in", module, "module"),
                   ylab = "Gene significance for Condition SvAll",
                   main = paste("Module membership vs. gene significance SvAll\n"),
                   cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)
```

Optional
```{r}
#=====================================================================================
#
#  Code chunk 6
#
#=====================================================================================
#Summary output of network analysis results
#Print all geneIDs in the analysis

#names(datExpr)
```

optional
```{r}
#=====================================================================================
#
#  Code chunk 7
#
#=====================================================================================
#Print all geneIDs in the the brown module


#names(datExpr)[moduleColors=="brown"]
```

match data to annotation file to connect gene IDs to gene names and other annotations
```{r}
#=====================================================================================
#
#  Code chunk 8
#
#=====================================================================================
# match data to annotation file to connect gene IDs to gene names and other annotations

annot = read.csv(file = "ref_assembly_corset.b2g.emapper.annot.csv");
dim(annot)
names(annot)
probes = names(datExpr)
probes2annot = match(probes, annot$GeneID)
# The following is the number or probes without annotation:
sum(is.na(probes2annot))
```

Create a data frame holding the following information for all probes in relation to Condition: 
```{r}
#=====================================================================================
#
#  Code chunk 9
#
#=====================================================================================
#Create a data frame holding the following information for all probes in relation to Condition: 
# probe ID, protein ID, gene name, module color, gene significance for weight, and module membership and p-values in all modules. 
#The modules will be ordered by their significance for weight, with the most significant ones to the left.

#Note: this will have to be done individually for each variable column

# Create the starting data frame
geneInfo0 = data.frame(GeneID = probes,
                       CorsetID = annot$CorsetID[probes2annot],
                       Description = annot$Description[probes2annot],
                       geneDescr = annot$description[probes2annot],
                       IPR = annot$InterPro.IDs[probes2annot],
                       geneName = annot$Preferred_name[probes2annot],
                       e.value = annot$e.value[probes2annot],
                       GO.IDs = annot$GO.IDs[probes2annot],
                       KEGG = annot$KEGG.KO[probes2annot],
                       KEGG.pathway = annot$KEGG.pathway[probes2annot],
                       Taxon = annot$best.tax.lvl[probes2annot],
                       moduleColor = moduleColors,
                       geneTraitSignificanceFvAll,
                       GSPvalueFvAll,
                       geneTraitSignificanceSvAll,
                       GSPvalueSvAll,
                       geneTraitSignificanceWvAll,
                       GSPvalueWvAll,
                       geneTraitSignificanceFS,
                       GSPvalueFS,
                       geneTraitSignificanceFW,
                       GSPvalueFW,
                       geneTraitSignificanceSW,
                       GSPvalueSW,
                       geneTraitSignificanceAQvAll,
                       GSPvalueAQvAll)

# Order modules by their significance for ConditionFvAll
modOrder = order(-abs(cor(MEs, ConditionFvAll, use = "p")));
# Add module membership information in the chosen order
for (mod in 1:ncol(geneModuleMembership))
{
  oldNames = names(geneInfo0)
  geneInfo0 = data.frame(geneInfo0, geneModuleMembership[, modOrder[mod]], 
                         MMPvalue[, modOrder[mod]]);
  names(geneInfo0) = c(oldNames, paste("MM.", modNames[modOrder[mod]], sep=""),
                       paste("p.MM.", modNames[modOrder[mod]], sep=""))
}
# Order the genes in the geneInfo variable first by module color, then by geneTraitSignificance
geneOrder = order(geneInfo0$moduleColor, -abs(geneInfo0$GS.ConditionFvAll));
geneInfo = geneInfo0[geneOrder, ]

```


Save data
```{r}
#=====================================================================================
#
#  Code chunk 10
#
#=====================================================================================

pl
write.csv(geneInfo, file = "WGCNA_4_Host-var20-geneInfo-Condition_13082024.csv")
```
