---
title: "WGCNA_2_data_input"
author: "BY"
date: "2024-08-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setup working directory and upload vst transformed counts
```{r}
#=====================================================================================
#
#  Code chunk 1
#
#=====================================================================================

#set working directory
setwd("~/Lucinid_project/BY04/WGCNA")
# Load the WGCNA package
library(WGCNA);
# The following setting is important, do not omit.
options(stringsAsFactors = FALSE);
#Read in the filtered host data set
HostData = read.csv("vstBY04_host_var20.csv");
# Take a quick look at what is in the data set:
dim(HostData); 
names(HostData);
#One can inspect larger data frames by invoking R data editor via fix(HostData)
```
Transpose the data
```{r}
#=====================================================================================
#
#  Code chunk 2
#
#=====================================================================================

#Remove auxiliary data and transpose expression data for further analysis.
datExpr0 = as.data.frame(t(HostData[, -c(1)]));
names(datExpr0) = HostData$gene_id;
rownames(datExpr0) = names(HostData)[-c(1)];
```

Check genes for samples with too many missing values
```{r}
#=====================================================================================
#
#  Code chunk 3
#
#=====================================================================================

#check for genes and samples with too many missing values
gsg = goodSamplesGenes(datExpr0, verbose = 3); 
#If this command returns TRUE, then all genes make the cut
gsg$allOK
```

Remove samples and genes with too many missing values if necessary
```{r}
#=====================================================================================
#
#  Code chunk 4
#
#=====================================================================================

#remove the offending genes and samples from the data:

if (!gsg$allOK)
{
  # Optionally, print the gene and sample names that were removed:
  if (sum(!gsg$goodGenes)>0) 
    printFlush(paste("Removing genes:", paste(names(datExpr0)[!gsg$goodGenes], collapse = ", ")));
  if (sum(!gsg$goodSamples)>0) 
    printFlush(paste("Removing samples:", paste(rownames(datExpr0)[!gsg$goodSamples], collapse = ", ")));
  # Remove the offending genes and samples from the data:
  datExpr0 = datExpr0[gsg$goodSamples, gsg$goodGenes]
}
```

cluster samples to identify obvious outliers
```{r}
#=====================================================================================
#
#  Code chunk 5
#
#=====================================================================================
#cluster samples to identify obvious outliers.

sampleTree = hclust(dist(datExpr0), method = "average");
# Plot the sample tree: Open a graphic output window of size 12 by 9 inches
# The user should change the dimensions if the window is too large or too small.
sizeGrWindow(12,9)
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, 
     cex.axis = 1.5, cex.main = 2)

pdf(file = "WGCNA_2_BY04_host-var20_dataInput_13082024_sampleclustering.pdf", width = 10, height = 8)

plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, 
     cex.axis = 1.5, cex.main = 2)
dev.off()

```
Optional step for removing outliers
```{r}
#=====================================================================================
#
#  Code chunk 6
#
#=====================================================================================

#If outliers exist, one can remove it by hand, or use an automatic approach. 
#Choose a height cut that will remove the offending sample, say 15 (the red line in the plot), and use a branch cut at that height.

# Plot a line to show the cut
#abline(h = 15, col = "red");
# Determine cluster under the line
#clust = cutreeStatic(sampleTree, cutHeight = 15, minSize = 10)
#table(clust)
# clust 1 contains the samples we want to keep.
#keepSamples = (clust==1)

#### There were no outliers so all samples are kept ####
datExpr = datExpr0
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)

#The variable datExpr now contains the expression data ready for network analysis.
```

Upload traits data
```{r}
#=====================================================================================
#
#  Code chunk 7
#
#=====================================================================================
#Read in the trait data and match the trait samples to the expression samples

traitData = read.csv("Traits.csv");
dim(traitData)
names(traitData)

# remove columns that hold information we do not need.
allTraits = traitData[, c(1, 3:11) ];
#allTraits = allTraits[, c(2, 11:36) ];
dim(allTraits)
names(allTraits)

# Form a data frame analogous to expression data that will hold the experimental traits.

Samples = rownames(datExpr);
traitRows = match(Samples, allTraits$Sample_ID);
datTraits = allTraits[traitRows, -1];
rownames(datTraits) = allTraits[traitRows, 1];

collectGarbage();

#We now have the expression data in the variable datExpr, and the corresponding clinical traits in the variable datTraits. 
```
We now have the expression data in the variable datExpr, and the corresponding clinical traits in the variable datTraits. 

Visualize how the experimental traits relate to the sample dendrogram
```{r}
#=====================================================================================
#
#  Code chunk 8
#
#=====================================================================================


#visualize how the experimental traits relate to the sample dendrogram.
# Re-cluster samples
sampleTree2 = hclust(dist(datExpr), method = "average")
# Convert traits to a color representation: white means low, red means high, grey means missing entry
traitColors = numbers2colors(datTraits, signed = FALSE);
# Plot the sample dendrogram and the colors underneath.
par(mar = c(6, 8, 4, 4));
plotDendroAndColors(sampleTree2, traitColors,
                    groupLabels = names(datTraits), 
                    main = "Sample dendrogram and trait heatmap")

pdf(file = "WGCNA_2_BY04_host-var20_dataInput_13082024_sampledendrogramvstraits.pdf", width = 10, height = 8)
par(mar = c(6, 8, 4, 4));
plotDendroAndColors(sampleTree2, traitColors,
                    groupLabels = names(datTraits), 
                    main = "Sample dendrogram and trait heatmap")
dev.off()
```
Save data for next step (WGCNA_3)
```{r}
#=====================================================================================
#
#  Code chunk 9
#
#=====================================================================================


save(datExpr, datTraits, file = "WGCNA_2_BY04-host-var20-dataInput_13082024.RData")

```
